---
title: "Reproducible Research Week 2 Assignment"
author: "Prachi Raut"
date: "6/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducible Research Week 2 Assignmanet
This assignment involves solving 6 questions based on analysis of data collected from an anonymus individual regarding number of steps taken.

## Loading and pre-processing the data

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
library(ggplot2)
library(scales)
library(Hmisc)
```


```{r, results='markup', warning=TRUE, message=TRUE}
activityData <- read.csv('activity.csv')
head(activityData)
```
## Transforming interval data

```{r}
#activityData$interval <- strptime(gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", activityData$interval), format='%H:%M')
```

-----

## 1. Total number of steps taken per day

```{r}
stepsByDay <- tapply(activityData$steps, activityData$date, sum, na.rm=TRUE)
```

## 2. Histogram of the total number of steps taken each day

```{r}
qplot(stepsByDay, xlab='Total steps per day', ylab='Frequency', binwidth = 500, color = I("black"), fill = I("white"))
```



## 2. Mean and median total number of steps taken per day

```{r}
stepsByDayMean <- mean(stepsByDay)
stepsByDayMedian <- median(stepsByDay)
```
* Mean: `r stepsByDayMean` 
* Median: `r stepsByDayMedian`

-----

## Average daily activity pattern

```{r}
averageStepsPerTimeBlock <- aggregate(x=list(meanSteps=activityData$steps), by=list(interval=activityData$interval), FUN=mean, na.rm=TRUE)
```

### Time series plot

```{r}
ggplot(data=averageStepsPerTimeBlock, aes(x=interval, y=meanSteps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken") 
```
### The 5-minute interval on average across all the days in the dataset that contains the maximum number of steps?

```{r}
mostSteps <- which.max(averageStepsPerTimeBlock$meanSteps)
timeMostSteps <-  gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", averageStepsPerTimeBlock[mostSteps,'interval'])
```

* Most Steps at: `r timeMostSteps`

----

## Imputing Missing values
### The total number of missing values in the dataset 

```{r}
numMissingValues <- length(which(is.na(activityData$steps)))
```

* Number of missing values: `r numMissingValues`

### Devise a strategy for filling in all of the missing values in the dataset.
The missing values are replaced by mean of the 5-minute interval. 

```{r}
activityDataImputed <- activityData
activityDataImputed$steps <- impute(activityData$steps, fun=mean)
```


### Histogram of the total number of steps taken each day 

```{r}
stepsByDayImputed <- tapply(activityDataImputed$steps, activityDataImputed$date, sum)
qplot(stepsByDayImputed, xlab='Total steps per day (Imputed)', ylab='Frequency', binwidth=500, col = I("black"), fill = I("white"))
```


 

### ... and Calculate and report the mean and median total number of steps taken per day. 

```{r}
stepsByDayMeanImputed <- mean(stepsByDayImputed)
stepsByDayMedianImputed <- median(stepsByDayImputed)
```
* Mean (Imputed): `r stepsByDayMeanImputed`
* Median (Imputed):  `r stepsByDayMedianImputed`


----

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels- weekday and weekend indicating whether a given date is a weekday or weekend day.


```{r}
activityDataImputed$dateType <-  ifelse(as.POSIXlt(activityDataImputed$date)$wday %in% c(0,6), 'weekend', 'weekday')
```

### Panel plot containing a time series plot


```{r}
averagedActivityDataImputed <- aggregate(steps ~ interval + dateType, data=activityDataImputed, mean)
ggplot(averagedActivityDataImputed, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(dateType ~ .) +
    xlab("5-minute interval") + 
    ylab("avarage number of steps")
```


